---
title: "BIEN Data for mapping and biostatistics"
output: html_notebook
---

[BIEN](http://biendata.org)

Today, to build on our previous lesson with BIEN, we will re-visit some of the work we did to be sure proper libraries and some data are loaded. (As you know, no need to re-install.)

```{r}
#install.packages("BIEN")
library("BIEN")
library(ape) #Package for working with phylogenies in R
library(maps) #Useful for making quick maps of occurrences
library(sp) # A package for spatial data
#vignette("BIEN")
```

```{r}
# Download the data for a focal species.
Artocarpus_altilis <- BIEN_occurrence_species(species = "Artocarpus altilis")
```

It looks like nothing happened, but you actually have created a data frame called Artocarpus_altilis, full of character data and numeric data. Use the commands below. The first allows you to view the so called meta-data (data about the data; how many observations, for example). The second command looks at the first six entries in the data frame. The third looks at the last six entries in the data frame.

```{r}
str(Artocarpus_altilis)
head(Artocarpus_altilis)
tail(Artocarpus_altilis)
```

As you know from looking at the browser, you can look up a rangemap. 
```{r}
breadfruit_rangemap <- BIEN_ranges_load_species("Artocarpus altilis")
```

```{r}
# Map the range of breadfruit
map('world', fill=T, col="grey", bg = "light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(breadfruit_rangemap, col="green", add=T)
```

```{r}
# Map the range of breadfruit + occurence records
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(breadfruit_rangemap, col="green", add=T)
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1)
```

By default, RBIEN commands for species occurrence only downloads limited information, but those defaults can be changed to get more expansive data, including:
- species in cultivation are added to those occurring naturally
- occurrence data outside of the Americas
- more taxonomy data (authority, higher levels than genus and species)
- when occurrence happened, country of occurrence, etc.

```{r}
Artocarpus_altilis_full <- BIEN_occurrence_species(species = "Artocarpus altilis",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Artocarpus_altilis_full)
```

This was a slightly slower process because so much more data were obtained. 

Now let's map again, but on the world map instead of just the Americas. And let's show points from the full query in red, and overlay points from the default query (more limited) in blue. 

```{r}
map('world', fill=T , col="grey", bg="light blue")

#Plot the points from the full query in red
points(cbind(Artocarpus_altilis_full$longitude, Artocarpus_altilis_full$latitude), col="red", pch=20, cex=1)

# Plot the points from the default query in blue
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1) 
```

### Trait Data and linking this to Biostatistics

```{r}
BIEN_trait_list()
```

```{r}
# Create data frame of breadfruit traits
breadfruit_traits <- BIEN_trait_genus(genus = "Artocarpus")
str(breadfruit_traits)
breadfruit_traits
```

```{r}
Breadfruit_DBH <- BIEN_trait_traitbygenus(trait = "diameter at breast height (1.3 m)", genus = "Artocarpus")
str(Breadfruit_DBH)
Breadfruit_DBH
```

```{r}
# List variable names from Breadfruit_DBH data frame
names(Breadfruit_DBH)
```

```{r}
Breadfruit_DBH_small = Breadfruit_DBH[c(1:5)]
Breadfruit_DBH_small
Breadfruit_DBH_small = as.data.frame(Breadfruit_DBH_small)
class(Breadfruit_DBH_small)
head(Breadfruit_DBH_small)
class(Breadfruit_DBH_small$trait_value)
DBH = as.numeric(Breadfruit_DBH_small$trait_value)
class(DBH)
```

Note that a library is being loaded below, and without it the boxplots will not be generated. This should be a package installed by default when R is installed or updated, but if it's not then you will need to install it.

```{r}
library(lattice)
head(Breadfruit_DBH_small)
boxplot(DBH~scrubbed_species_binomial, data=Breadfruit_DBH_small)
```
  	
```{r}
Breadfruit_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Artocarpus")
str(Breadfruit_WPH)
Breadfruit_WPH
```

```{r}
names(Breadfruit_WPH)
```

```{r}
Breadfruit_WPH_small = Breadfruit_WPH[c(1:5)]
Breadfruit_WPH_small
Breadfruit_WPH_small = as.data.frame(Breadfruit_WPH_small)
class(Breadfruit_WPH_small)
head(Breadfruit_WPH_small)
class(Breadfruit_WPH_small$trait_value)
WPH = as.numeric(Breadfruit_WPH_small$trait_value)
class(WPH)
```

```{r}
library(lattice)
head(Breadfruit_WPH_small)
hist(WPH)
boxplot(WPH~scrubbed_species_binomial, data=Breadfruit_WPH_small)
```

```{r}
mean(DBH, data=Breadfruit_DBH_small)
mean(WPH, data=Breadfruit_WPH_small)
```

Below there is another package being installed into the library and then loaded from the library to be active in the R software. This allows summarizing within species (rather than the whole genus), generating standard functions such as means and standard deviations (sd). The online resource Quick-R can be helpful if you want a bit more support in generating summary statistics, especially within sub-groups. 

https://www.statmethods.net/stats/descriptives.html

```{r}
#install(doBy)
library(doBy)
summaryBy(DBH~scrubbed_species_binomial, data=Breadfruit_DBH_small, FUN=c(mean, sd))
summaryBy(WPH~scrubbed_species_binomial, data=Breadfruit_WPH_small, FUN=c(mean, sd))
```

```{r}
(unique(Breadfruit_DBH_small$scrubbed_species_binomial))
```

```{r}
(unique(Breadfruit_WPH_small$scrubbed_species_binomial))
```

## Exercises for completing the portfolio 
Do you think the difference in whole plant height is statistically significant between Artocarpus hirsutus and Artocarpus integer?

Can you combine what was learned in the previous lesson "Introduction to getting, testing and plotting data"? 

Can you choose the most appropriate of the following two tests, re-write and run the code, and create the last portion of this R Notebook to include output, and some interpretive comments about statistical support for or against that hypothesis? 

Also try this for a second species pair, which require obtaining some data from BIEN using R BIEN. 

Some possibilities might be: 

cashews are from Anacardium occidentale and perhaps one of the wild cashews (see Wikipedia)

avocados are from Persea americana but there are other relatives in the genus Persea, notably P. coyo

chocolate and one (or a group) of its relatives within Theobroma

choose two species in the genus Agave (used for fiber, tequila; see Wikipedia)

# Population Mean Between Two Independent Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-independent-samples

# Population Mean Between Two Matched Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-matched-samples

```{r}
# Population Mean Between Two Independent Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-independent-samples
ttest_result1 = t.test(x=Breadfruit_WPH_small$scrubbed_species_binomial, y=Breadfruit_WPH_small$trait_value, paired=FALSE, alternative="two.sided")
ttest_result1
```

# This will prepare you for Digital Lab 2

```{r}
#BIEN_occurrence_country()
bronx <- BIEN_occurrence_county(country="United States", state="New York", county = "Bronx", native.status=TRUE)
bronx
```

# Hilary add stuff here!
